A semantics of core AUTOSAR

### 2016-08-29 paper/paper.tex
Paper: merge
Nuked the port meta-variables!

Components and ports do not carry any dynamic state besides the state of their constituent parts, so they need no explicit representation as process terms. However, in order to still be able to talk about model elements using their locally scoped names, we employ a hierarchical addressing scheme, such that address $\Tr{\V{R}}{\V{I}}$ means runnable \V{R} of component \V{I} and $\Tep{\V{E}}{\V{P}}{\V{I}}$ means element \V{E} of port \V{P} within component \V{I}. We use the following meta-variable conventions:

%
\[
\begin{array}{rcll}
  \V{I} & \in & \text{Component names} \\
  \V{R} & \in & \text{Runnable names} \\
  \V{P} & \in & \text{Port names} \\
  \V{E} & \in & \text{Sender-receiver element names} \\
  \V{O} & \in & \text{Client-server operation names} \\
  \V{S} & \in & \text{Inter-runnable variable names} \\
  \V{X} & \in & \text{Exclusive area names} \\
\\
  \V{A},\V{B} & \in & \text{Addresses} \\
  \V{A},\V{B} & ::= & \Tr{\V{R}}{\V{I}}   \sep   i.p  \sep  \Tep{\V{E}}{\V{P}}{\V{I}}   \sep \Top{\V{O}}{\V{P}}{\V{I}}   \sep   \Ts{\V{S}}{\V{I}}   \sep   \Tx{\V{X}}{\V{I}}   \\
\end{array}
\]
%
Since AUTOSAR components can be arbitrarily nested, the component names \V{I} can also be considered to have a similarly nested internal structure (component \V{I_1} within component \V{I_2} within component \V{I_3}, etc). We can fully ignore this detail here, though, as the hierarchical layout of components in an AUTOSAR system has no run-time implications \cite[ch.~3.3.1]{AR:SWC}.

The process terms of our calculus are as follows:
\begin{itemize}
\item $\Trunnable {\Tr r i} t \Act n$  \newline
Dynamic state for runnable $r$ of component $i$, indicating $n$ current instances, $t$ seconds left until next possible instantiation, and activation state $\Act$.
\item $\Trinst {\Tr r i} c \Xs \Code$ \newline
An instance of runnable $r$ of component $i$, currently executing $\Code$ and owning the exclusive areas $\Xs$, possibly running on behalf of client $c$.
\item $\Tirv {\Ts s i} v$ \newline
Inter-runnable variable $s$ of component $i$, currently holding value $v$.
\item $\Texcl {\Tx x i} u$ \newline
Exclusive area $x$ of component $i$, with boolean occupation state $u$.
\item $\Tdelem {\Tep e p i} u v$ \newline
Non-queued sender-receiver data element $e$ in port $p$ of component $i$, currently holding value $v$ with boolean update status flag $u$.
\item $\Tqelem {\Tep e p i} n \Vs$ \newline
Queued sender-receiver data element $e$ in port $p$ of component $i$ with maximum capacity $n$, currently holding value sequence $\Vs$.
\item $\Toper {\Top o p i} m \Srv$ \newline
Client-server operation $o$ in port $p$ of component $i$, currently holding sequence number $m$ and call state $\Srv$.
\item $\Ttimer {\Tr r i} {t_p} t$ \newline
Periodic timer for runnable $r$ of component $i$, with a period of $t_p$ and $t$ seconds left of its current cycle.
\end{itemize}

In the untyped setting of this report, a value $v$ can be any data item computed and communicated by an AUTOSAR system, including the unit value $\Tvoid$ and the error codes that might be returned from RTE calls. Meta-variables $n$ and $m$ stand for natural numbers, while $u$ and $t$ range over boolean values and (floating point) time values, respectively. The client $c$ of a runnable instance is a tuple holding the address, sequence number and argument value of the current invocation (if the runnable was invoked by a client-server call), or is $\Tvoid$ otherwise. An activation state $\Act$ either toggles between $\Tidle$ and $\Tpending$, or is a sequence of client tuples (in the case of a runnable triggered by client-server calls). A call state $\Srv$ alternates between indicating an ongoing call with a timeout and a finished call with a result value.
Formally,
\[
\begin{array}{rcll}
  v        & \in & \text{Values} \\
  n,m      & \in & \text{Natural numbers} \\
  u        & \in & \text{Boolean values} \\
  t        & \in & \text{Time values} \\
  c        & \in & \text{Clients} \\
  \Act     & \in & \text{Activation states} \\
  \Srv     & \in & \text{Call states} \\ \\
  c        & ::= & \Tclient {\Top o p i} m v \sep \Tvoid \\
  \Act     & ::= & \Tidle \sep \Tpending \sep \Cs \\
  \Srv     & ::= & \Tcalling t \sep \Tdone v
\end{array}
\]
We write $\Vs$ for a sequence of values \V{V}, $\Cs$ for a sequence of clients \V{C}, etc. Sequences may be empty, which is written~$\epsilon$. By $\Tlength{\Vs}$ we mean the length of sequence $\Vs$. Left (right) concatenation of an element with a sequence is written $\Tcons{\V{V}}{\V{Vs}}$ ($\Tsnoc{\V{Vs}}{\V{V}}$).

@@

To this end, we have chosen to ignore all internal computations and just capture the observable effects of runnable execution as a pure sequence of RTE (VFB) calls. And because the result from an RTE call may in general affect subsequent runnable behaviour, we use a continuation-passing style where every $\Code$ term except the final return command literally contains the sequence that follows it, as a continuation parameter $\Cont$.
%
\[
\begin{array}{lll}
  \Code & \in & \text{Code}                \\
  \Cont & \in & \text{Values} \rightarrow \text{Code}   \\ \\
  \Code & ::= & \Enter x \Cont               \\
        & |   & \Exit x \Cont                \\
        & |   & \IrvWrite s v \Cont          \\
        & |   & \IrvRead s \Cont             \\
        & |   & \Send {\Te e p} v \Cont      \\
        & |   & \Receive {\Te e p} \Cont     \\
        & |   & \Write {\Te e p} v \Cont     \\
        & |   & \Read {\Te e p} \Cont        \\
        & |   & \Invalidate {\Te e p} \Cont  \\
        & |   & \IsUpdated {\Te e p} \Cont   \\
        & |   & \Call {\To o p} v \Cont      \\
        & |   & \Result {\To o p} \Cont      \\
        & |   & \Return v                    \\
\end{array}
\]

@@

The label $l$ of a transition can essentially be of two kinds, indicating that the affected process either ``says'' or ``hears'' something during the transition. Such a label always consists of an address $a$ paired with some additional detail $d$. There is also a special form of ``hearing'' that denotes letting time pass and which does not carry any address. The different labels are captured in the following grammar.
\[
\begin{array}{rcll}
  l         & ::= & a!d       & \hspace{4em}\text{Say  \V{A} and \V{D}}   \\
            & |   & a?d       & \hspace{4em}\text{Hear \V{A} and \V{D}}   \\
            & |   & \Tdelta t & \hspace{4em}\text{Let \V{T} seconds pass} \\
\end{array}
\]
The possible forms of the extra label information \V{D} will be introduced as the different transition rules are defined.

@@

\item For each data element $e$ of each required port $p$ of each atomic component $i$, a term
\[
\begin{array}{c}
  \Tqelem {\Tep e p i} n \epsilon
\end{array}
\]
if the \emph{swImplPolicy} attribute of element $e$ is \emph{queued} (with capacity $n$); otherwise, a term
\[
\begin{array}{c}
  \Tdelem {\Tep e p i} \Tfalse \TneverReceived
\end{array}
\]

\item For each operation $o$ of each required port $p$ of each atomic component $i$, a term
\[
\begin{array}{c}
  \Toper{\Top o p i} 0 {\Tdone \Tvoid}
\end{array}
\]

\item For each \emph{TimingEvent} of each runnable $r$ of each atomic component $i$, a term
\[
\begin{array}{c}
  \Ttimer {\Tr r i } {t_p} 0
\end{array}
\]
where $\V{T_p}$ is the period of the event.

@@

Some further static model information will also be referenced by the transition rules. We refer to the connectors of a model using a binary relation $\Tconnect{}{}$ on port addresses, where the arrow points in the \emph{provided-to-required} direction. We also assume that $\Tconnect{}{}$ is transitively closed (i.e., it expresses end-to-end connectivity across \emph{Delegation-} as well as \emph{AssemblySwConnector}s), and that every connection $\Tconnect{a}{b}$ is lifted to the \emph{DataElement}s and \emph{ClientServerOperation}s of the connected ports such that
\[
\begin{array}{lll}
  \Tconnect{a.e}{b.e} & \text{for all elements $e$ common to $a$ and $b$} \\
  \Tconnect{a.o}{b.o} & \text{for all operations $o$ common to $a$ and $b$} \\
\end{array}
\]

@@

We write $\Timplementation{i.r}{k}$ to state the assumption that the static AUTOSAR model under study assigns implementation function $k$ to runnable $i.r$. Similar notations will be used to express other references to the underlying static model; for example $\TdataReceivedEvent {i.r} {i.p.e}$ to state that the model allows runnable $i.r$ to be triggered by data reception events on port $i.p.e$. Although the chosen notation is sometimes significantly shorter than the UML/XML-based syntax used in the AUTOSAR specification, the intended meaning should nevertheless be clear.

@@

Axioms (13) and (15) define unbuffered inter-process communication from a runnable instance point of view. These transitions are just syntactic variants of shared variable reading and writing (axioms (5) and (7)). A data element (\textbf{delem}) process term contains two pieces of state in addition to its address: a communicated data value and a boolean flag for keeping track of unread writes. Axioms (14) and (16) capture reads and writes to the data value and also set the update flag accordingly. The side-condition in axiom (16) makes the rule apply only if there is a static connection from the data element being written and the address of the matched term (recall that data element storage is associated with the receiving side of a connection). Because AUTOSAR allows sender-receiver communication patterns to be one-to-many (such as $\Tconnect {i.p.e} a$ and $\Tconnect {i.p.e} b$), parallel data element updates like the following are possible:
$$
\begin{array}{c}
  \Trinst {i.r} c \Xs {\TrteWrite {p.e} v k} \,\Opar  \\
  {\Tdelem a \_ \_} \Opar {\Tdelem b \_ \_} \vspace{0.6ex} \\
  \red{\say{i.p.e}{\Twr v}} \\
  \Trinst {i.r} c \Xs {\Tap k \Tvoid} \,\Opar \\
  {\Tdelem a \Ttrue v} \Opar {\Tdelem b \Ttrue v}
\end{array}
$$
As an aside, note that the ordering of terms in a parallel composition is made entirely irrelevant by the symmetric formulation of transition rules (1) and (2).

Axiom (17) provides another means of interpreting a data element write. It expresses that a runnable set up to trigger on data reception on required data element $a$ should be marked as \textbf{pending} whenever a write occurs on an element connected to $a$. An example may be as follows (assuming $\Tconnect {i.p.e} a$ and $\TdataReceivedEvent b a$:
$$
\begin{array}{c}
  \Trinst {i.r} c \Xs {\TrteWrite {p.e} v k} \,\Opar  \\
  {\Tdelem a \_ \_} \Opar {\Trunnable b t \_ n} \vspace{0.6ex} \\
  \red{\say{i.p.e}{\Twr v}} \\
  \Trinst {i.r} c \Xs {\Tap k \Tvoid} \,\Opar \\
  {\Tdelem a \Ttrue v} \Opar {\Trunnable b t \Tpending n}
\end{array}
$$
Transitions that actually create an instance of a pending runnable will be introduced in Section~\ref{sect:SpawnTerm}.

@@

The following example illustrates a scenario where a sent value is only stored in a subset of the connected buffers. Assuming $\Tconnect {\Tep e p i} a$ and $\Tconnect {\Tep e p i} b$:
$$
\begin{array}{c}
  \Trinst {i.r} c \Xs {\TrteSend {p.e} v k} \,\Opar  \\
  {\Tqelem a 2 \Teps} \Opar {\Tqelem b 2 {\Tcons {v_1} {v_2}}} \vspace{0.6ex} \\
  \red{\say{i.p.e}{\Tsnd v b}} \\
  \Trinst {i.r} c \Xs {\Tap k \Tlimit} \,\Opar \\
  {\Tqelem a 2 v} \Opar {\Tqelem b 2 {\Tcons {v_1} {v_2}}}
\end{array}
$$

@@

Here is an example of how a client runnable instance, a client port operation, and a server runnable interact during a call transition (assuming $\ToperationInvokedEvent b a$, $\Tconnect a {i.p.o}$, and $\TserverCallPointTimeout{i.p.o} t $).
$$
\begin{array}{c}
  {\Trinst {i.r} c \Xs {\TrteCall {p.o} v k}} \,\Opar  \\
  {\Toper {i.p.o} m {\Tdone \Tnodata}} \Opar \\
  {\Trunnable b 0 \Cs 0} \vspace{0.6ex} \\
  \red{\say{i.p.o}{\Tcall m v}} \\
  {\Trinst {i.r} c \Xs {\Tap k \Tok}} \,\Opar \\
  {\Toper {i.p.o} m {\Tcalling t}} \Opar \\
  {\Trunnable b 0 {\Tcons \Cs {\Tclient {i.p.o} m v}}  0}
\end{array}
$$

@@

An example of a normal server result interaction can look as follows:
$$
\begin{array}{c}
  {\Trinst b {\Tclient {i.p.o} {91} {v_0}} \Xs {\Treturn v}} \,\Opar  \\
  {\Toper {i.p.o} {91} {\Tcalling t}} \vspace{0.6ex} \\
  \red{\say{i.p.o}{\Tret {91} v}} \\
  {\Trinst b \Tvoid \Xs {\Treturn \Tvoid}} \,\Opar \\
  {\Toper {i.p.o} {92} {\Tdone v}}
\end{array}
$$
As a contrast, this is the behaviour observed when a call timeout occurs:
$$
\begin{array}{c}
  {\Trinst b {\Tclient {i.p.o} {91} {v_0}} \Xs {\Treturn v}} \,\Opar  \\
  {\Toper {i.p.o} {91} {\Tcalling 0}} \vspace{0.6ex} \\
  \red{\say{i.p.o}{\Tret {91} \Ttimeout}} \\
  {\Trinst b {\Tclient {i.p.o} {91} {v_0}} \Xs {\Treturn v}} \,\Opar  \\
  {\Toper {i.p.o} {92} {\Tdone \Ttimeout}}
\end{array}
$$
The processes are now in a state where runnable instance $b$ is prohibited from doing a normal return. Progress is only possible via axioms (46) and (47).
$$
\begin{array}{c}
  {\Trinst b {\Tclient {i.p.o} {91} {v_0}} \Xs {\Treturn v}} \,\Opar  \\
  {\Toper {i.p.o} {92} {\Tdone \Ttimeout}} \vspace{0.6ex} \\
  \red{\say{i.p.o}{\Tskip {91}}} \\
  {\Trinst b \Tvoid \Xs {\Treturn \Tvoid}} \,\Opar \\
  {\Toper {i.p.o} {92} {\Tdone \Ttimeout}}
\end{array}
$$

@@

Complementing the notion of spawning is a mechanism for runnable instance termination. An instance with a $\Tvoid$ client parameter (either acquired natively, or via axioms (42) or (46)) transforms into the terminated process $\Tzero$ as defined in axiom (51). Since this term is the unit of parallel process composition, it is silently absorbed by any other process in parallel with it. The instance termination is also noted by the associated runnable parent, which reacts by decrementing its current instance counter (axiom (52)). An example of this interaction follows below.
$$
\begin{array}{c}
  {\Trunnable b t \Act 3} \Opar
  {\Trinst b \Tvoid \Teps {\Treturn v}} \vspace{0.6ex} \\
  \red{\say{i.p.o}{\Tterm}} \\
  {\Trunnable b t \Act 2} \Opar \Tzero \\
  \equiv \\
  {\Trunnable b t \Act 2}
\end{array}
$$

@@

As an concrete illustration, consider the example of Section~\ref{sect:BufSndRcv}. The axioms that enable each individual process transition in that example are (26), (27) and (28), respectively (and rules (1) and (3) allow them to be combined). In particular, axiom (28) is applicable because of the assumption $\Tconnect {i.p.e} b$, which simultaneously blocks applicability of axiom (70). However, under the different assumption that there is no connection from ${i.p.e}$ to $b$, axioms (28) and (70) must trade places. This also forces the removal of $b$ from the second argument of the \payload{snd} payload (i.e., the list of connected buffers not able to store the value sent). As a consequence, axiom (26) is no longer applicable and must be replaced by axiom (25). The resulting parallel transition looks as follows, illustrating the correct generation of an \Tatom{ok} result despite the presence of a full (but unconnected) buffer.
$$
\begin{array}{c}
  \Trinst {i.r} c \Xs {\TrteSend {p.e} v k} \,\Opar  \\
  {\Tqelem a 2 \Teps} \Opar {\Tqelem b 2 {\Tcons {v_1} {v_2}}} \vspace{0.6ex} \\
  \red{\say{i.p.e}{\Tsnd v \Teps}} \\
  \Trinst {i.r} c \Xs {\Tap k \Tok} \,\Opar \\
  {\Tqelem a 2 v} \Opar {\Tqelem b 2 {\Tcons {v_1} {v_2}}}
\end{array}
$$

@@

The AUTOSAR specification contains a very subtle sensitivity to this kind of non-deterministism that is worth examplifying. Consider a queued data element $a$ together with a runnable $b$ set up to trigger on data reception events for $a$. A scenario where two values $v_1$ and $v_2$ are sent on a port element $i.p.e$ connected to $a$ may lead to the following trace:
$$
\begin{array}{c}
  {\Tqelem a 2 \Teps} \Opar
  {\Trunnable b 0 \Tidle 0} \vspace{0.6ex} \\

  \red{\hear{i.p.e}{\Tsnd {v_1} \Teps}} \\

@@

  \red{\hear{i.p.e}{\Tsnd {v_2} \Teps}} \\

@@

  \red{\hear{i.p.e}{\Tsnd {v_1} \Teps}} \\

@@

Unqueued data elements that are read before first written should return the error code {\bf neverReceived} \cite[ch.~5.6.10]{AR:RTE}. Since no exceptions to this rule are mentioned in the specification, our intepretation is that this holds even when a data element possesses an \emph{initValue} attribute  (initial {\bf delem} state in Section~\ref{sec:Sem}). However, as this renders any such initial value assignments entirely redundant, an alternative interpretation would be to create a term
$$
  \Tdelem {i.p.e} \Tfalse v
$$
for each unqueued data element $i.p.e$, where $v$ is the \emph{initValue} of $e$ if it exists, and $\TneverReceived$ otherwise.

>>>>

Components and ports do not carry any dynamic state besides the state of their constituent parts, so they need no explicit representation as process terms. However, in order to still be able to talk about model elements using their locally scoped names, we employ a hierarchical addressing scheme, such that address $i.r$ means runnable $r$ of component $i$ and $i.e$ means port element $e$ within component $i$. We use the following meta-variable conventions:

%
\[
\begin{array}{rcll}
  i & \in & \text{Component names} \\
  r & \in & \text{Runnable names} \\
  e & \in & \text{Sender-receiver port element names} \\
  o & \in & \text{Client-server port operation names} \\
  s & \in & \text{Inter-runnable variable names} \\
  x & \in & \text{Exclusive area names} \\
\\
  a,b & \in & \text{Addresses} \\
  a,b & ::= & i.r  \sep  i.e   \sep i.o   \sep   i.s   \sep   i.x   \\
\end{array}
\]
%
AUTOSAR elements and operations are actually only unique within the {\em ports} they belong to, so technically we ought to address them by reference to an intended {\em port name} as well, as in $i.p.e$ or $i.p.o$ for some port $p$. However, port names play no role of their own in our study, so we choose to leave them be implied by the element and operation names we reference. Moreover, component names \V{I} can also be considered to have a nested internal structure (component \V{I_1} within component \V{I_2} within component \V{I_3}, etc). We can fully ignore this detail here, though, as the hierarchical layout of components in an AUTOSAR system has no run-time implications \cite[ch.~3.3.1]{AR:SWC}.

The process terms of our calculus are as follows:
\begin{itemize}
\item $\Trunnable {i.r} t \Act n$  \newline
Dynamic state for runnable $r$ of component $i$, indicating $n$ current instances, $t$ seconds left until next possible instantiation, and activation state $\Act$.
\item $\Trinst {i.r} c \Xs \Code$ \newline
An instance of runnable $r$ of component $i$, currently executing $\Code$ and owning the exclusive areas $\Xs$, possibly running on behalf of client $c$.
\item $\Tirv {i.s} v$ \newline
Inter-runnable variable $s$ of component $i$, currently holding value $v$.
\item $\Texcl {i.x} u$ \newline
Exclusive area $x$ of component $i$, with boolean occupation state $u$.
\item $\Tdelem {i.e} u v$ \newline
Non-queued sender-receiver port element $e$ of component $i$, currently holding value $v$ with boolean update status flag $u$.
\item $\Tqelem {i.e} n \Vs$ \newline
Queued sender-receiver port element $e$ of component $i$ with maximum capacity $n$, currently holding value sequence $\Vs$.
\item $\Toper {i.o} m \Srv$ \newline
Client-server port operation $o$ of component $i$, currently holding sequence number $m$ and call state $\Srv$.
\item $\Ttimer {i.r} {t_p} t$ \newline
Periodic timer for runnable $r$ of component $i$, with a period of $t_p$ and $t$ seconds left of its current cycle.
\end{itemize}

In the untyped setting of this report, a value $v$ can be any data item computed and communicated by an AUTOSAR system, including the unit value $\Tvoid$ and the error codes that might be returned from RTE calls. Meta-variables $n$ and $m$ stand for natural numbers, while $u$ and $t$ range over boolean values and (floating point) time values, respectively. The client $c$ of a runnable instance is a tuple holding the address, sequence number and argument value of the current invocation (if the runnable was invoked by a client-server call), or is $\Tvoid$ otherwise. An activation state $\Act$ either toggles between $\Tidle$ and $\Tpending$, or is a sequence of client tuples (in the case of a runnable triggered by client-server calls). A call state $\Srv$ alternates between indicating an ongoing call with a timeout and a finished call with a result value.
Formally,
\[
\begin{array}{rcll}
  v        & \in & \text{Values} \\
  n,m      & \in & \text{Natural numbers} \\
  u        & \in & \text{Boolean values} \\
  t        & \in & \text{Time values} \\
  c        & \in & \text{Clients} \\
  \Act     & \in & \text{Activation states} \\
  \Srv     & \in & \text{Call states} \\ \\
  c        & ::= & \Tclient {i.o} m v \sep \Tvoid \\
  \Act     & ::= & \Tidle \sep \Tpending \sep \Cs \\
  \Srv     & ::= & \Tcalling t \sep \Tdone v
\end{array}
\]
We write $\Vs$ for a sequence of values \V{V}, $\Cs$ for a sequence of clients \V{C}, etc. Sequences may be empty, which is written~$\epsilon$. By $\Tlength{\Vs}$ we mean the length of sequence $\Vs$. Left (right) concatenation of an element with a sequence is written $\Tcons{\V{V}}{\V{Vs}}$ ($\Tsnoc{\V{Vs}}{\V{V}}$).

@@

To this end, we have chosen to ignore all internal computations and just capture the observable effects of runnable execution as a pure sequence of RTE (VFB) calls. And because the result from an RTE call may in general affect subsequent runnable behaviour, we use a continuation-passing style where every $\Code$ term except the final return command literally contains the sequence that follows it, as a continuation parameter $\Cont$.
%
\[
\begin{array}{lll}
  \Code & \in & \text{Code}                \\
  \Cont & \in & \text{Values} \rightarrow \text{Code}   \\ \\
  \Code & ::= & \Enter x \Cont               \\
        & |   & \Exit x \Cont                \\
        & |   & \IrvWrite s v \Cont          \\
        & |   & \IrvRead s \Cont             \\
        & |   & \Send e v \Cont      \\
        & |   & \Receive e \Cont     \\
        & |   & \Write e v \Cont     \\
        & |   & \Read e \Cont        \\
        & |   & \Invalidate e \Cont  \\
        & |   & \IsUpdated e \Cont   \\
        & |   & \Call o v \Cont      \\
        & |   & \Result o \Cont      \\
        & |   & \Return v                    \\
\end{array}
\]

@@

The label $l$ of a transition can essentially be of two kinds, indicating that the affected process either ``says'' or ``hears'' something during the transition. Such a label always consists of an address $a$ paired with some additional detail $d$. There is also a special form of ``hearing'' that denotes letting time pass and which does not carry any address. The different labels are captured in the following grammar.
\[
\begin{array}{rcll}
  l         & ::= & a!d       & \hspace{4em}\text{Say  \V{A} and \V{D}}   \\
            & |   & a?d       & \hspace{4em}\text{Hear \V{A} and \V{D}}   \\
            & |   & \Tdelta t & \hspace{4em}\text{Let \V{T} seconds pass} \\
\end{array}
\]
The possible forms of the extra label detail \V{D} will be introduced as the different transition rules are defined.

@@

\item For each {\em required} port element $e$ of each atomic component $i$, a term
\[
\begin{array}{c}
  \Tqelem {i.e} n \epsilon
\end{array}
\]
if the \emph{swImplPolicy} attribute of element $e$ is \emph{queued} (with capacity $n$); otherwise, a term
\[
\begin{array}{c}
  \Tdelem {i.e} \Tfalse \TneverReceived
\end{array}
\]

\item For each {\em required} port operation $o$ of each atomic component $i$, a term
\[
\begin{array}{c}
  \Toper{i.o} 0 {\Tdone \Tvoid}
\end{array}
\]

\item For each \emph{TimingEvent} of each runnable $r$ of each atomic component $i$, a term
\[
\begin{array}{c}
  \Ttimer {i.r} {t_p} 0
\end{array}
\]
where $\V{T_p}$ is the period of the event.

@@

Some further static model information will also be referenced by the transition rules. We refer to the connectors of a model using a binary relation $\Tconnect{}{}$ on port elements and operations, where the arrow points in the \emph{provided-to-required} direction. Although connectors are tecnically defined for ports, we assume $\Tconnect{}{}$ is lifted to the common elements and operations of connected ports, and that it is transitively closed (i.e., expresses end-to-end connectivity across \emph{Delegation-} as well as \emph{AssemblySwConnector}s).

@@

We write $\Timplementation{i.r}{k}$ to state the assumption that the static AUTOSAR model under study assigns implementation function $k$ to runnable $i.r$. Similar notations will be used to express other references to the underlying static model; for example $\TdataReceivedEvent {i.r} {i.e}$ to state that the model allows runnable $i.r$ to be triggered by data reception events on port element $i.e$. Although the chosen notation is sometimes significantly shorter than the UML/XML-based syntax used in the AUTOSAR specification, the intended meaning should nevertheless be clear.

@@

Axioms (13) and (15) define unbuffered inter-process communication from a runnable instance point of view. These transitions are just syntactic variants of shared variable reading and writing (axioms (5) and (7)). A data element (\textbf{delem}) process term contains two pieces of state in addition to its address: a communicated data value and a boolean flag for keeping track of unread writes. Axioms (14) and (16) capture reads and writes to the data value and also set the update flag accordingly. The side-condition in axiom (16) makes the rule apply only if there is a static connection from the data element being written and the address of the matched term (recall that data element storage is associated with the receiving side of a connection). Because AUTOSAR allows sender-receiver communication patterns to be one-to-many (such as $\Tconnect {i.e} a$ and $\Tconnect {i.e} b$), parallel data element updates like the following are possible:
$$
\begin{array}{c}
  \Trinst {i.r} c \Xs {\TrteWrite {p.e} v k} \,\Opar  \\
  {\Tdelem a \_ \_} \Opar {\Tdelem b \_ \_} \vspace{0.6ex} \\
  \red{\say{i.e}{\Twr v}} \\
  \Trinst {i.r} c \Xs {\Tap k \Tvoid} \,\Opar \\
  {\Tdelem a \Ttrue v} \Opar {\Tdelem b \Ttrue v}
\end{array}
$$
As an aside, note that the ordering of terms in a parallel composition is made entirely irrelevant by the symmetric formulation of transition rules (1) and (2).

Axiom (17) provides another means of interpreting a data element write. It expresses that a runnable set up to trigger on data reception on required data element $a$ should be marked as \textbf{pending} whenever a write occurs on an element connected to $a$. An example may be as follows (assuming $\Tconnect {i.e} a$ and $\TdataReceivedEvent b a$:
$$
\begin{array}{c}
  \Trinst {i.r} c \Xs {\TrteWrite {p.e} v k} \,\Opar  \\
  {\Tdelem a \_ \_} \Opar {\Trunnable b t \_ n} \vspace{0.6ex} \\
  \red{\say{i.e}{\Twr v}} \\
  \Trinst {i.r} c \Xs {\Tap k \Tvoid} \,\Opar \\
  {\Tdelem a \Ttrue v} \Opar {\Trunnable b t \Tpending n}
\end{array}
$$
Transitions that actually create an instance of a pending runnable will be introduced in Section~\ref{sect:SpawnTerm}.

@@

The following example illustrates a scenario where a sent value is only stored in a subset of the connected buffers. Assuming $\Tconnect {i.e} a$ and $\Tconnect {i.e} b$:
$$
\begin{array}{c}
  \Trinst {i.r} c \Xs {\TrteSend {p.e} v k} \,\Opar  \\
  {\Tqelem a 2 \Teps} \Opar {\Tqelem b 2 {\Tcons {v_1} {v_2}}} \vspace{0.6ex} \\
  \red{\say{i.e}{\Tsnd v b}} \\
  \Trinst {i.r} c \Xs {\Tap k \Tlimit} \,\Opar \\
  {\Tqelem a 2 v} \Opar {\Tqelem b 2 {\Tcons {v_1} {v_2}}}
\end{array}
$$

@@

Here is an example of how a client runnable instance, a client port operation, and a server runnable interact during a call transition (assuming $\ToperationInvokedEvent b a$, $\Tconnect a {i.o}$, and $\TserverCallPointTimeout{i.o} t $).
$$
\begin{array}{c}
  {\Trinst {i.r} c \Xs {\TrteCall {p.o} v k}} \,\Opar  \\
  {\Toper {i.o} m {\Tdone \Tnodata}} \Opar \\
  {\Trunnable b 0 \Cs 0} \vspace{0.6ex} \\
  \red{\say{i.o}{\Tcall m v}} \\
  {\Trinst {i.r} c \Xs {\Tap k \Tok}} \,\Opar \\
  {\Toper {i.o} m {\Tcalling t}} \Opar \\
  {\Trunnable b 0 {\Tcons \Cs {\Tclient {i.o} m v}}  0}
\end{array}
$$

@@

An example of a normal server result interaction can look as follows:
$$
\begin{array}{c}
  {\Trinst b {\Tclient {i.o} {91} {v_0}} \Xs {\Treturn v}} \,\Opar  \\
  {\Toper {i.o} {91} {\Tcalling t}} \vspace{0.6ex} \\
  \red{\say{i.o}{\Tret {91} v}} \\
  {\Trinst b \Tvoid \Xs {\Treturn \Tvoid}} \,\Opar \\
  {\Toper {i.o} {92} {\Tdone v}}
\end{array}
$$
As a contrast, this is the behaviour observed when a call timeout occurs:
$$
\begin{array}{c}
  {\Trinst b {\Tclient {i.o} {91} {v_0}} \Xs \Code} \,\Opar  \\
  {\Toper {i.o} {91} {\Tcalling 0}} \vspace{0.6ex} \\
  \red{\say{i.o}{\Tret {91} \Ttimeout}} \\
  {\Trinst b {\Tclient {i.o} {91} {v_0}} \Xs \Code} \,\Opar  \\
  {\Toper {i.o} {92} {\Tdone \Ttimeout}}
\end{array}
$$
The processes are now in a state where runnable instance $b$ is prohibited from doing a normal return. Termination of $b$ is only possible via axioms (46) and (47).
$$
\begin{array}{c}
  {\Trinst b {\Tclient {i.o} {91} {v_0}} \Xs {\Treturn v}} \,\Opar  \\
  {\Toper {i.o} {92} {\Tdone \Ttimeout}} \vspace{0.6ex} \\
  \red{\say{i.o}{\Tskip {91}}} \\
  {\Trinst b \Tvoid \Xs {\Treturn \Tvoid}} \,\Opar \\
  {\Toper {i.o} {92} {\Tdone \Ttimeout}}
\end{array}
$$

@@

Complementing the notion of spawning is a mechanism for runnable instance termination. An instance with a $\Tvoid$ client parameter (either acquired natively, or via axioms (42) or (46)) transforms into the terminated process $\Tzero$ as defined in axiom (51). Since this term is the unit of parallel process composition, it is silently absorbed by any other process in parallel with it. The instance termination is also noted by the associated runnable parent, which reacts by decrementing its current instance counter (axiom (52)). An example of this interaction follows below.
$$
\begin{array}{c}
  {\Trunnable b t \Act 3} \Opar
  {\Trinst b \Tvoid \Teps {\Treturn v}} \vspace{0.6ex} \\
  \red{\say{i.o}{\Tterm}} \\
  {\Trunnable b t \Act 2} \Opar \Tzero \\
  \equiv \\
  {\Trunnable b t \Act 2}
\end{array}
$$

@@

As an concrete illustration, consider the example of Section~\ref{sect:BufSndRcv}. The axioms that enable each individual process transition in that example are (26), (27) and (28), respectively (and rules (1) and (3) allow them to be combined). In particular, axiom (28) is applicable because of the assumption $\Tconnect {i.e} b$, which simultaneously blocks applicability of axiom (70). However, under the different assumption that there is no connection from ${i.e}$ to $b$, axioms (28) and (70) must trade places. This also forces the removal of $b$ from the second argument of the \payload{snd} payload (i.e., the list of connected buffers not able to store the value sent). As a consequence, axiom (26) is no longer applicable and must be replaced by axiom (25). The resulting parallel transition looks as follows, illustrating the correct generation of an \Tatom{ok} result despite the presence of a full (but unconnected) buffer.
$$
\begin{array}{c}
  \Trinst {i.r} c \Xs {\TrteSend {p.e} v k} \,\Opar  \\
  {\Tqelem a 2 \Teps} \Opar {\Tqelem b 2 {\Tcons {v_1} {v_2}}} \vspace{0.6ex} \\
  \red{\say{i.e}{\Tsnd v \Teps}} \\
  \Trinst {i.r} c \Xs {\Tap k \Tok} \,\Opar \\
  {\Tqelem a 2 v} \Opar {\Tqelem b 2 {\Tcons {v_1} {v_2}}}
\end{array}
$$

@@

The AUTOSAR specification contains a very subtle sensitivity to this kind of non-deterministism that is worth examplifying. Consider a queued data element $a$ together with a runnable $b$ set up to trigger on data reception events for $a$. A scenario where two values $v_1$ and $v_2$ are sent on a port element $i.e$ connected to $a$ may lead to the following trace:
$$
\begin{array}{c}
  {\Tqelem a 2 \Teps} \Opar
  {\Trunnable b 0 \Tidle 0} \vspace{0.6ex} \\

  \red{\hear{i.e}{\Tsnd {v_1} \Teps}} \\

@@

  \red{\hear{i.e}{\Tsnd {v_2} \Teps}} \\

@@

  \red{\hear{i.e}{\Tsnd {v_1} \Teps}} \\

@@

Unqueued data elements that are read before first written should return the error code {\bf neverReceived} \cite[ch.~5.6.10]{AR:RTE}. Since no exceptions to this rule are mentioned in the specification, our intepretation is that this holds even when a data element possesses an \emph{initValue} attribute  (initial {\bf delem} state in Section~\ref{sec:Sem}). However, as this renders any such initial value assignments entirely redundant, an alternative interpretation would be to create a term
$$
  \Tdelem {i.e} \Tfalse v
$$
for each unqueued data element $i.e$, where $v$ is the \emph{initValue} of $e$ if it exists, and $\TneverReceived$ otherwise.
